Stability Analysis of Recurrent Neural Networks \\ using Ordinary Differential Equations

### 2020-07-06 p5/stability_rnn.tex
First review

\section{Notation}\label{s2}
We follow the contemporary mathematical practice of not using separate
notation for vectors and scalars. Instead, we define every symbol.

@@

The oldest method to solve equation \eqref{s3e1} numerically was devised 
by Euler in the eighteenth century. It follows straight from the definition
of the first derivative of a function. If we know the value of $y$ at $x_0$,
we can approximate the value of $y$ at a neighbouring point $x_0 + h$ as
\begin{equation}\label{s3e2}
y(x_0 + h) = y(x_0) + h\left[\frac{dy}{dx}\right]_{x_0}.
\end{equation}
But the value of the derivative of $y$ with respect to $x$ is given by
the differential equation itself. Thus,
\begin{equation}\label{s3e3}
y(x_0 + h) = y(x_0) + h f(x_0, y(x_0)).
\end{equation}
In general, $n$ steps beyond the first, one can write
\begin{equation}\label{s3e4}
y(x_n + h) = y(x_{n}) + h f(x_{n}, y(x_{n})).
\end{equation}
Euler's algorithm is not the best numerical algorithm to solve ODE's but
it provides the germ for the the rest.

\section{RNN with a Single Layer}\label{s4}
Recurrent neural networks, abbreviated as RNN, feed a part of their output
to their input. The output if a step $n$ thus depends on both the input
at that step and the output at the previous step. This makes them similar
to the ordinary differential equations described in the previous section.
Suppose we have a single layer neural network that is updated using the
following rule:
\begin{equation}\label{s4e1}
y_{n+1} = \sigma(w_x x_{n} + w_y y_n),
\end{equation}
where $w_x$ and $w_y$ are weights and $\sigma$ is an activation function. 
If we now identify the correspondence
\begin{eqnarray}
y_{n+1} &\rightarrow& y(x_n + h) \\
y_{n} &\rightarrow& y(x_n) \\
\sigma(w_x x_n + w_y y_n) &\rightarrow& y(x_n) + hf(x_n, y(x_n))
\end{eqnarray}
then we can spot a striking resemblance between Euler's algorithm and an
RNN. The idea that an RNN corresponds to an ODE and therefore the stability
of an RNN can be studied using the stability of the underlying ODE is the
topic of this article.

We need not interpret equation \eqref{s3e1} as a scalar equation. We can
as well let $y$ and $f$ be vector functions. That is, $y:\mathbb{R}^p
\mapsto \mathbb{R}^q$, $x \in \mathbb{R}^p$ and $f: \mathbb{R}^{p+q}
\mapsto \mathbb{R}^q$. In this case, equation \eqref{s3e4} becomes
\begin{equation}\label{s4e5}
y(x_n + h) = y(x_n) + \nabla y(x_n)\cdot h.
\end{equation}
Here
\begin{equation}\label{s4e6}
\nabla y = \begin{pmatrix} y_{1, x_1} & \ldots & y_{1, x_p} \\
	   \vdots \\
	   y_{q, x_1} & \ldots & y_{q, x_p}
	   \end{pmatrix}
\end{equation}
is an $q \times p$ matrix and $h \in \mathbb{R}^p$ so that their product
$\nabla y(x_n) \cdot h$ is in $\mathbb{R}^q$. An RNN corresponding to this
equation is
\begin{equation}\label{s4e7}
y_{n+1} = \sigma\left(W_x x_n + W_y y_n\right),
\end{equation}
where $y_n$ and $y_{n+1}$ are $q$-dimensional vectors, $x_n \in \mathbb{R}
^p$, $W_x$ is an $q \times p$ weight matrix for the input and $W_y$ is an
$q \times q$ weight matrix for the output.

\section{Higher order ODE}\label{s5}
A system of first order ODE is a set of equations of the form
\begin{eqnarray*}
\frac{dy_1}{dx} &=& f_1(x, y_1, \ldots, y_k) \\
\vdots \\
\frac{dy_k}{dx} &=& f_k(x, y_1, \ldots, y_k) \\
\end{eqnarray*}
It can be written in a vector form as
\begin{equation}\label{s5e1}
\frac{dy}{dx} = f,
\end{equation}
where
\[
y = \begin{pmatrix}y_1 \\ \vdots \\ y_k\end{pmatrix},\;\text{and}\;
f = \begin{pmatrix}f_1 \\ \vdots \\ f_k\end{pmatrix}
\]
A system of first order ODE is interesting because they are equivalent to
a higher order differential equation. For example, one can write the $k$th
order ODE
\begin{equation}\label{s5e2}
\frac{d^ky}{dx^k} = f\left(x, y, \frac{dy}{dx}, \ldots, 
	\frac{d^{k-1}y}{dx^{k-1}}\right)
\end{equation}
as the system
\begin{eqnarray*}
\frac{dy_{k-1}}{dx} &=& f(x, y, y_1, \ldots, y_{k-1}) \\
\frac{dy_{k-2}}{dx} &=& y_{k-1} \\
\vdots \\
\frac{dy}{dx} &=& y_1
\end{eqnarray*}
If this equation is also accompanied by an initial value then we can solve
it numerically using equations \eqref{s4e5}, \eqref{s4e6} and \eqref{s4e7}.

\section{Runge-Kutta methods}\label{s6}
Runge-Kutta methods are a collection of numerical algorithms developed by
the German mathematicians Carl Runge and Martin Kutta in the late 
nineteenth centuries\cite{lambert1991numerical}. We will describe the most
commonly used among them, called the `RK4' method. If $y: \mathbb{R}
\mapsto \mathbb{R}$ is the unknown function that obeys the ODE
\begin{equation}\label{s6e1}
\frac{dy}{dt} = f(t, y), y(t_0) = y_0
\end{equation}
then one can get the value oy $y$ at $t_n + h$ using the formula
\begin{equation}\label{s6e2}
y(t_n + h) = y(t_n) + \frac{h}{6}(k_1 + 2k_2 + 2k_3 + k_4).
\end{equation}
The constants $k_1, \ldots, k_4$ are given by
\begin{eqnarray}
k_1 &=& f(t_n, y(t_n)) \label{s6e3} \\
k_2 &=& f\left(t_n + \frac{h}{2}, y(t_n) + h\frac{k_1}{2}\right) 
 \label{s6e4} \\
k_3 &=& f\left(t_n + \frac{h}{2}, y(t_n) + h\frac{k_2}{2}\right) 
 \label{s6e5} \\
k_4 &=& f(t_n + h, y_n + hk_3). \label{s6e6}
\end{eqnarray}
Let us study these equations carefully. The value of $y$ at $t_n + h$ 
depends on $y(t_n)$ and four other numbers $k_1, \ldots, k_4$. Of these,
$k_1$ depends on $y(t_n)$ and $t_n$, $k_2$ on $y(t_n)$, $t_n$ and $k_1$,
$k_3$ on $y(t_n), t_n$ and $k_2$ and $k_4$ on $y(t_n), t_n$ and $k_3$. 
Runge-Kutta methods perform much better then Euler methods when $f$ is 
a non-linear function of $x$. The number of intermediate values, $4$ in 
this case, is called the \emph{the number of stages} of the algorithm.

\section{RNN with Hidden Layers}\label{s7}
Let us try to interpret the `RK4' algorithm as a neural network. It is 
clearly a recurrent neural network because the output at `time' $t_n + h$
depends on that at `time' $t_n$. If we write $y(t_n) = y_n$ then we can
say that $y_{n+1}$ depends on $y_n$. However, it does not depend on $y_n$
alone but on four other numbers which must be computed sequentially. $k_1$
is computed from $t_n$ and $y_n$. The rest from $t_n$, $y_n$ and the 
`previous' $k$. We can thus construct a neural network with four hidden 
layers, each computing $k_1$ to $k_4$ and the output later combining
all of them to get the final result. This is a $4$-stage, $1$st order
RNN.

We now define a general RNN\cite{niu2019recurrent}.
\begin{defn}\label{s7d1}
An ODE-RNN of $n$th order in non-linearity and $t$th order in gradient,
called $(n,t)$-ODERNN, where $n, t \ge 1$ are integers is described by
the following update rule.
\begin{eqnarray}
K_1 &=& \sigma_1(W_1Y_l + b_1) \label{s7e1} \\
K_q &=& \gamma_q K_{q-1} + \kappa_1\sigma_q\left(
 W_qY_{l-t_{q-1}}+b_q+h\sum_{k=1}^{q-1}\alpha_{q,k}K_k\right)\label{s7e2}\\
Y_{l+1} &=& \gamma_{n+1}Y_l + \kappa_{n+1}\sigma_{q+1}\left(
 W_{n+1}Y_{l-t_{n-1}}+b_{n+1}+h\sum_{k=1}^n\beta_kK_k
 \right)\label{s7e3},
\end{eqnarray}
where
\begin{enumerate}
\item $q = 1, \ldots, n$. This explains the $n$ in the $(n,t)$-ODERNN.
\item The right hand side of equations \eqref{s7e1} to \eqref{s7e3} 
depend on $Y_l$ to $Y_{l - t_{n-1}}$, a total of $n$ previous values.
Here each $t_k = \lfloor tk/n\rfloor$. Because $t_k$ for each hidden layer
is defined in terms of $t$, we also specify it in the name of the ODE. 
This explain $t$ in $(n, t)$-ODERNN.
\item $Y_l \in \mathbb{R}^s$, $l \in \mathbb{Z}$;
\item $W_1,\ldots, W_{n+1}$ are $p \times s$ weight matrices;
\item $b_1,\ldots, b_{n+1} \in \mathbb{R}^p$ are bias vectors;
\item $\alpha_{q, k}, \beta_k$, for $q = 2, \ldots, n$ and $k = 1, \ldots,
n$, $\gamma_2, \ldots, \gamma_{n}, \kappa_2, \ldots, \kappa_{n}$ are 
$p \times p$ matrices;
\item $\gamma_{n+1}$ is an $s \times s$ matrix and $\kappa_{n+1}$ is a 
$s \times p$ matrix.
\item $h \in \mathbb{R}$ is a constant and
\item $\sigma_1, \ldots, \sigma_{q+1}$ are activation function defined as
\[
\sigma_i\begin{pmatrix}x_1 \\ \vdots \\ x_p\end{pmatrix} = 
\begin{pmatrix}\sigma_i(x_1) \\ \vdots \\ \sigma_i(x_p)\end{pmatrix},
i = 1, \ldots, q+1.
\]
\end{enumerate}
\end{defn}
\noindent Before proceeding, a few remarks are in order.
\begin{rem}
$K_1, \ldots, K_q$ are all $p$-dimensional vectors. That is, they 
are all members of $\mathbb{R}^p$.
\end{rem}
\begin{rem}
We can readily confirm that the arguments to the activation functions
$\sigma_1, \ldots, \sigma_{q+1}$ are all $p$-dimensional vectors.
\end{rem}
\begin{rem}
The number of hidden layers is $n$. The RNN thus has $n + 2$ layers. 
\end{rem}

>>>>

\section{Notation}\label{s2}
We follow the contemporary mathematical practice of not using separate
notation for vectors and scalars. Instead, we define every symbol. Whenver
a symbol stands for a vector, it denotes a column vector.

@@

The oldest method to solve equation \eqref{s3e1} numerically was devised 
by Euler in the eighteenth century. It follows straight from the definition
of the first derivative of a function. If we know the value of $y$ at $x_0$,
we can approximate the value of $y$ at a neighbouring point $x_0 + h$ as
\begin{equation}\label{s3e2}
y(x_0 + h) = y(x_0) + h\left[\frac{dy}{dx}\right]_{x_0}.
\end{equation}
But the value of the derivative of $y$ with respect to $x$ is given by
the differential equation itself. Thus,
\begin{equation}\label{s3e3}
y(x_0 + h) = y(x_0) + h f(x_0, y(x_0)).
\end{equation}
In general, $n$ steps beyond the first, one can write
\begin{equation}\label{s3e4}
y(x_n + h) = y(x_{n}) + h f(x_{n}, y(x_{n})).
\end{equation}
Euler's algorithm is not the best numerical algorithm to solve ODE's but
it provides the germ for the rest.

\section{RNN with a Single Layer}\label{s4}
Recurrent neural networks, abbreviated as RNN, feed a part of their output
to their input. The output of a step $n$ thus depends on both the input
at that step and the output of the previous step. This makes them similar
to the ordinary differential equations described in the previous section.
Suppose we have a single layer neural network that is updated using the
following rule:
\begin{equation}\label{s4e1}
y_{n+1} = \sigma(w_x x_{n} + w_y y_n),
\end{equation}
where $w_x$ and $w_y$ are weights and $\sigma$ is an activation function. 
If we now identify the correspondence
\begin{eqnarray}
y_{n+1} &\rightarrow& y(x_n + h) \\
y_{n} &\rightarrow& y(x_n) \\
\sigma(w_x x_n + w_y y_n) &\rightarrow& y(x_n) + hf(x_n, y(x_n))
\end{eqnarray}
then we can spot a striking resemblance between Euler's algorithm and an
RNN. The idea that an RNN corresponds to an ODE and therefore the stability
of an RNN can be studied using the stability of the underlying ODE is the
topic of this article.

We need not interpret equation \eqref{s3e1} as a scalar equation. We can
as well let $y$ and $f$ be vector functions. That is, $y:\mathbb{R}^p
\mapsto \mathbb{R}^q$, $x \in \mathbb{R}^p$ and $f: \mathbb{R}^{p+q}
\mapsto \mathbb{R}^q$. In this case, equation \eqref{s3e4} becomes
\begin{equation}\label{s4e5}
y(x_n + h) = y(x_n) + \nabla y(x_n)\cdot h.
\end{equation}
Here
\begin{equation}\label{s4e6}
\nabla y = \begin{pmatrix} y_{1, x_1} & \ldots & y_{1, x_p} \\
	   \vdots \\
	   y_{q, x_1} & \ldots & y_{q, x_p}
	   \end{pmatrix}
\end{equation}
is an $q \times p$ matrix and $h \in \mathbb{R}^p$ so that their product
$\nabla y(x_n) \cdot h$ is in $\mathbb{R}^q$. The symbol $y_{i, x_j}$ 
stands for the partial derivative of the $i$th component of $y$, denoted by
$y_i$ with respect to $x_j$. Thus,
\[
    y_{i, x_j} = \frac{\partial y_i}{\partial x_j}.
\]
An RNN corresponding to this equation is
\begin{equation}\label{s4e7}
y_{n+1} = \sigma\left(W_x x_n + W_y y_n\right),
\end{equation}
where $y_n$ and $y_{n+1}$ are $q$-dimensional vectors, $x_n \in \mathbb{R}
^p$, $W_x$ is an $q \times p$ weight matrix for the input and $W_y$ is an
$q \times q$ weight matrix for the output.

\section{Higher order ODE}\label{s5}
We can interpret the vector form of equation \eqref{s3e1} as a system of 
first order ODE
\begin{eqnarray*}
\frac{dy_1}{dx} &=& f_1(x, y_1, \ldots, y_k) \\
\vdots \\
\frac{dy_k}{dx} &=& f_k(x, y_1, \ldots, y_k) \\
\end{eqnarray*}
Its equivalent vector form is
\begin{equation}\label{s5e1}
\frac{d}{dx}\begin{pmatrix}y_1 \\ \vdots \\ y_k\end{pmatrix} = 
\begin{pmatrix}f_1 \\ \vdots \\ f_k\end{pmatrix}.
\end{equation}
A system of first order ODE is interesting because it is equivalent to
a higher order differential equation. For example, one can write the $k$th
order ODE
\begin{equation}\label{s5e2}
\frac{d^ky}{dx^k} = f\left(x, y, \frac{dy}{dx}, \ldots, 
	\frac{d^{k-1}y}{dx^{k-1}}\right)
\end{equation}
as the system
\begin{eqnarray*}
\frac{dy_{k-1}}{dx} &=& f(x, y, y_1, \ldots, y_{k-1}) \\
\frac{dy_{k-2}}{dx} &=& y_{k-1} \\
\vdots \\
\frac{dy}{dx} &=& y_1
\end{eqnarray*}
If this equation is also accompanied by an initial value then we can solve
it numerically using equations \eqref{s4e5}, \eqref{s4e6} and \eqref{s4e7}.

\section{Runge-Kutta methods}\label{s6}
Runge-Kutta methods are a collection of numerical algorithms developed by
the German mathematicians Carl Runge and Martin Kutta in the late 
nineteenth century\cite{lambert1991numerical}. We will describe the most
commonly used among them, called the `RK4' method. If $y: \mathbb{R}
\mapsto \mathbb{R}$ is the unknown function that obeys the ODE
\begin{equation}\label{s6e1}
\frac{dy}{dt} = f(t, y), y(t_0) = y_0
\end{equation}
then one can get the value oy $y$ at $t_n + h$ using the formula
\begin{equation}\label{s6e2}
y(t_n + h) = y(t_n) + \frac{h}{6}(k_1 + 2k_2 + 2k_3 + k_4).
\end{equation}
The constants $k_1, \ldots, k_4$ are given by
\begin{eqnarray}
k_1 &=& f(t_n, y(t_n)) \label{s6e3} \\
k_2 &=& f\left(t_n + \frac{h}{2}, y(t_n) + h\frac{k_1}{2}\right) 
 \label{s6e4} \\
k_3 &=& f\left(t_n + \frac{h}{2}, y(t_n) + h\frac{k_2}{2}\right) 
 \label{s6e5} \\
k_4 &=& f(t_n + h, y_n + hk_3). \label{s6e6}
\end{eqnarray}
Let us study these equations carefully. The value of $y$ at $t_n + h$ 
depends on $y(t_n)$ and four other numbers $k_1, \ldots, k_4$. Each of
these is the value of the derivative of $y$ at different points. Of these,
$k_1$ depends on $y(t_n)$ and $t_n$, $k_2$ on $y(t_n)$, $t_n$ and $k_1$,
$k_3$ on $y(t_n), t_n$ and $k_2$ and $k_4$ on $y(t_n), t_n$ and $k_3$. If
one were to implement this algorithm, every iteration of it would consist
of using the output of the previous iteration and a calculation of four
constants, one after another, to get the value of the current iteration.
This is indeed like a recurrent neural network with four hidden layers.

Runge-Kutta methods perform much better then Euler methods when $f$ is 
a non-linear function of $x$. The number of intermediate values, $4$ in 
this case, is called the \emph{the number of stages} of the algorithm.

\section{RNN with Hidden Layers}\label{s7}
Let us try to interpret the `RK4' algorithm as a neural network. It is 
clearly a recurrent neural network because the output at `time' $t_n + h$
depends on that at `time' $t_n$. If we write $y(t_n) = y_n$ then we can
say that $y_{n+1}$ depends on $y_n$. However, it does not depend on $y_n$
alone but on four other numbers which must be computed sequentially. $k_1$
is computed from $t_n$ and $y_n$. The rest from $t_n$, $y_n$ and the 
`previous' $k$. We can thus construct a neural network with four hidden 
layers, each computing $k_1$ to $k_4$ and the output later combining
all of them to get the final result. We call a recurrent neural network
implementing an algorithm to solve an ordinary differential equation an
ODERNN.

We now define a general ODERNN\cite{niu2019recurrent}.
\begin{defn}\label{s7d1}
An ODE-RNN of $n$-th order in non-linearity and $t$-th order in gradient,
called $(n,t)$-ODERNN, where $n, t \ge 1$ are integers is described by
the following update rule.
\begin{eqnarray}
K_1 &=& \sigma_1(W_1Y_l + b_1) \label{s7e1} \\
K_q &=& \gamma_q K_{q-1} + \kappa_1\sigma_q\left(
 W_qY_{l-t_{q-1}}+b_q+h\sum_{k=1}^{q-1}\alpha_{q,k}K_k\right)\label{s7e2}\\
Y_{l+1} &=& \gamma_{n+1}Y_l + \kappa_{n+1}\sigma_{q+1}\left(
 W_{n+1}Y_{l-t_{n-1}}+b_{n+1}+h\sum_{k=1}^n\beta_kK_k
 \right)\label{s7e3},
\end{eqnarray}
where
\begin{enumerate}
\item $q = 1, \ldots, n$. This explains the $n$ in the $(n,t)$-ODERNN.
\item The right hand side of equations \eqref{s7e1} to \eqref{s7e3} 
depend on $Y_l$ to $Y_{l - t_{n-1}}$, a total of $n$ previous values.
Here each $t_k = \lfloor tk/n\rfloor$. Because $t_k$ for each hidden layer
is defined in terms of $t$, we also specify it in the name of the ODE. 
This explain $t$ in $(n, t)$-ODERNN.
\item $Y_l \in \mathbb{R}^s$, $l \in \mathbb{Z}$;
\item $W_1,\ldots, W_{n+1}$ are $p \times s$ weight matrices;
\item $b_1,\ldots, b_{n+1} \in \mathbb{R}^p$ are bias vectors;
\item $\alpha_{q, k}, \beta_k$, for $q = 2, \ldots, n$ and $k = 1, \ldots,
n$, $\gamma_2, \ldots, \gamma_{n}, \kappa_2, \ldots, \kappa_{n}$ are 
$p \times p$ matrices;
\item $\gamma_{n+1}$ is an $s \times s$ matrix and $\kappa_{n+1}$ is a 
$s \times p$ matrix.
\item $h \in \mathbb{R}$ is a constant and
\item $\sigma_1, \ldots, \sigma_{q+1}$ are activation function defined as
\[
\sigma_i\begin{pmatrix}x_1 \\ \vdots \\ x_p\end{pmatrix} = 
\begin{pmatrix}\sigma_i(x_1) \\ \vdots \\ \sigma_i(x_p)\end{pmatrix},
i = 1, \ldots, q+1.
\]
\end{enumerate}
\end{defn}
\noindent Before proceeding, a few remarks are in order.
\begin{rem}
$K_1, \ldots, K_q$ are all $p$-dimensional vectors. That is, they 
are all members of $\mathbb{R}^p$.
\end{rem}
\begin{rem}
We can readily confirm that the arguments to the activation functions
$\sigma_1, \ldots, \sigma_{q+1}$ are all $p$-dimensional vectors.
\end{rem}
\begin{rem}
The number of hidden layers is $n$. The RNN thus has $n + 2$ layers. 
\end{rem}
